---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stanr

<!-- badges: start -->
<!-- badges: end -->

`stanr` is an R interface for compiling Stan programs and running Stan
services directly from R. It uses an R6-based API similar to `cmdstanr`:
`stan_model()` returns a `StanModel` object, and inference methods
(`$sample()`, `$optimize()`, `$variational()`, etc.) return R6 fit objects
with methods for extracting draws, summaries, and diagnostics. Results use
`posterior` draw objects where possible.

## Inference Methods

| Method | Purpose | Return type |
| --- | --- | --- |
| `$sample()` | MCMC sampling (HMC/NUTS, static HMC, fixed param) | `StanMCMC` |
| `$optimize()` | Posterior mode or maximum likelihood estimate | `StanMLE` |
| `$laplace()` | Laplace approximation draws around a mode | `StanLaplace` |
| `$variational()` | ADVI approximate posterior draws | `StanVB` |
| `$pathfinder()` | Pathfinder approximate posterior draws | `StanPathfinder` |
| `$generate_quantities()` | Generated quantities from existing draws | `StanGQ` |
| `$diagnose()` | Gradient checking diagnostic | `StanDiagnose` |

## Fit Object Methods

All fit objects inherit from `StanFit` and share common methods:

| Method | Description |
| --- | --- |
| `$draws()` | Extract draws as `posterior` objects |
| `$summary()` | Summarize draws via `posterior::summarise_draws()` |
| `$print()` | Print a summary table |
| `$return_codes()` | Stan return codes (0 = success) |
| `$metadata()` | Fit metadata (seed, data, arguments) |
| `$time()` | Timing information |
| `$log_prob()` | Evaluate log probability |
| `$constrain_variables()` | Constrain unconstrained parameters |
| `$unconstrain_variables()` | Unconstrain parameters |
| `$save_object()` | Save fit to file |

## Installation

You can install the development version of stanr from GitHub:

``` r
# install.packages("pak")
options(pkg.git_submodules = TRUE)
pak::pak("git::https://github.com/andrjohns/stanr")
```

## Compile a Model

Models are compiled from a string or from a `.stan` file. `stan_model()`
returns a `StanModel` R6 object.

```{r compile-model}
library(stanr)

bernoulli_model <- "
data {
  int<lower=0> N;
  array[N] int<lower=0, upper=1> y;
}
parameters {
  real<lower=0, upper=1> theta;
}
model {
  theta ~ beta(1, 1);
  y ~ bernoulli(theta);
}
generated quantities {
  array[N] real log_lik;
  for (n in 1:N) {
    log_lik[n] = bernoulli_lpmf(y[n] | theta);
  }
}
"

mod <- stan_model(code = bernoulli_model, model_name = "bernoulli")
data <- list(N = 10, y = c(1, 0, 1, 1, 0, 1, 0, 0, 1, 0))
```

## Sampling

Call `$sample()` on a `StanModel` to run MCMC. The result is a
`StanMCMC` object with methods for extracting draws and diagnostics.

```{r sampling}
fit <- mod$sample(
  data = data,
  iter_warmup = 20,
  iter_sampling = 20,
  chains = 2,
  seed = 123,
  show_messages = FALSE
)

posterior::variables(fit$draws())
```

Extract a summary via `$summary()`:

```{r sampling-summary}
fit$summary()
```

Access sampler diagnostics and chain information:

```{r sampling-diagnostics}
list(
  chains = fit$num_chains(),
  return_codes = fit$return_codes(),
  diagnostics = fit$diagnostic_summary()
)
```

### Fixed Param

For models without parameters, use `fixed_param = TRUE`:

```{r fixed-param}
fixed_param_model <- "
data {
  int<lower=0> N;
}
generated quantities {
  array[N] int<lower=0, upper=1> y_rep;
  for (n in 1:N) {
    y_rep[n] = bernoulli_rng(0.5);
  }
}
"

fixed_mod <- stan_model(code = fixed_param_model, model_name = "fixed_param")

fixed_fit <- fixed_mod$sample(
  data = list(N = 5),
  iter_warmup = 0,
  iter_sampling = 10,
  fixed_param = TRUE,
  seed = 123,
  show_messages = FALSE
)

posterior::variables(fixed_fit$draws())
```

### Supplying Inverse Metrics

`$sample()` accepts precomputed inverse metric values through `inv_metric`.

For a diagonal metric, pass a numeric vector with one value per unconstrained
parameter:

```{r sampling-diag-metric}
diag_fit <- mod$sample(
  data = data,
  metric = "diag_e",
  inv_metric = c(1.0),
  iter_warmup = 20,
  iter_sampling = 20,
  seed = 123,
  show_messages = FALSE
)

posterior::variables(diag_fit$draws())
```

For a dense metric, pass a square matrix. To provide one metric per chain,
pass a list whose length equals `chains`:

```{r sampling-dense-metric}
dense_fit <- mod$sample(
  data = data,
  metric = "dense_e",
  inv_metric = list(matrix(1.0, nrow = 1, ncol = 1)),
  chains = 1,
  iter_warmup = 20,
  iter_sampling = 20,
  seed = 123,
  show_messages = FALSE
)

posterior::variables(dense_fit$draws())
```

## Pathfinder

Call `$pathfinder()` for an approximate posterior based on L-BFGS paths.

```{r pathfinder}
pf <- mod$pathfinder(
  data = data,
  max_lbfgs_iters = 100,
  num_paths = 1,
  draws = 50,
  seed = 123,
  show_messages = FALSE
)

posterior::variables(pf$draws())
pf$summary()
```

## Variational Inference

Call `$variational()` for ADVI with either a mean-field or full-rank Gaussian
family.

```{r variational}
vb <- mod$variational(
  data = data,
  algorithm = "meanfield",
  iter = 1000,
  draws = 50,
  seed = 123,
  show_messages = FALSE
)

posterior::variables(vb$draws())
vb$summary()
```

## Optimization

Call `$optimize()` to find a posterior mode or maximum likelihood estimate.

```{r optimizing}
opt <- mod$optimize(
  data = data,
  algorithm = "lbfgs",
  seed = 123,
  show_messages = FALSE
)

opt$mle()
opt$summary()
```

## Laplace Approximation

Call `$laplace()` to draw from a Gaussian approximation around a mode. Pass a
`StanMLE` object or let `$laplace()` run optimization first.

```{r laplace}
lap <- mod$laplace(
  data = data,
  mode = opt,
  draws = 20,
  seed = 123,
  show_messages = FALSE
)

posterior::variables(lap$draws())
lap$summary()
```

## Generated Quantities

Call `$generate_quantities()` with a `StanFit` object or draws matrix to
evaluate the generated quantities block.

```{r generated-quantities}
gq <- mod$generate_quantities(
  fitted_params = fit,
  data = data,
  seed = 123,
  show_messages = FALSE
)

posterior::variables(gq$draws())
gq$summary()
```

## Gradient Check

Call `$diagnose()` to compare autodiff gradients with finite-difference
approximations.

```{r gradient-check}
diag <- mod$diagnose(
  data = data,
  seed = 123
)

list(
  num_failed = diag$num_failed(),
  lp = diag$lp(),
  gradients = diag$gradients()
)
```

## Tuples and Complex Numbers

Stan `tuple(...)` values map to/from **unnamed** R lists (one element per
slot; arrays of tuples become lists of such lists), and `complex` /
`complex_vector` / `complex_row_vector` / `complex_matrix` map to/from R's
native complex type -- for `data =`, `init =`, draws, and exposed functions
alike.

```{r tuple-complex}
tuple_complex_code <- "
functions {
  tuple(real, vector) split_stat(vector x) {
    return (mean(x), head(x, 2));
  }
  complex_vector rotate(complex_vector z, complex phase) {
    return z * phase;
  }
}
"

tc_mod <- stan_model(
  code = tuple_complex_code,
  model_name = "tuple_complex",
  compile = FALSE
)
tc_mod$expose_stan_functions()

# tuple(real, vector) comes back as an unnamed list: the scalar slot, then
# the vector slot.
tc_mod$functions$split_stat(c(1, 2, 3, 4))

# complex_vector arguments/returns round-trip as native R complex vectors.
tc_mod$functions$rotate(c(1 + 2i, 3 - 1i), 1i)
```

## Stan Compilation Helpers

`stanc()` compiles Stan code to C++ and returns the generated C++ source as a
single string.

```{r stanc}
cpp_code <- stanc(bernoulli_model)
substr(cpp_code, 1, 80)
```

`stan_model()` is the higher-level entry point: it calls `stanc()`, compiles
the generated C++, and returns a `StanModel` object used by all inference
methods.
