---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# newstan

<!-- badges: start -->
<!-- badges: end -->

`newstan` is an R interface for compiling Stan programs and running Stan
services directly from R. Results use `posterior` draw objects where possible,
and every service result includes the arguments passed through to the backend.

## Service Return Types

| Function | Purpose | Return type |
| --- | --- | --- |
| `sampling()` | MCMC sampling, including HMC/NUTS and fixed-parameter runs | list with `draws`, `diagnostics`, `return_code`, `args` |
| `pathfinder()` | Pathfinder approximate posterior draws | list with `draws`, `diagnostics`, `return_code`, `args` |
| `variational()` | ADVI approximate posterior draws | list with `draws`, `return_code`, `args` |
| `optimizing()` | Posterior mode or maximum likelihood estimate | list with `par`, `value`, `return_code`, `args` |
| `laplace()` | Laplace approximation draws around a mode | list with `draws`, `return_code`, `args` |
| `generated_quantities()` | Generated quantities from existing parameter draws | list with `draws`, `return_code`, `args` |
| `gradient_check()` | Finite-difference gradient diagnostic | integer count of failed checks |

## Installation

You can install the development version of newstan from GitHub:

``` r
# install.packages("pak")
pak::pak("andrjohns/newstan")
```

## Compile A Model

Models can be compiled from a string or from a `.stan` file. The returned object
is a model environment used by the service functions below.

```{r compile-model}
library(newstan)

bernoulli_model <- "
data {
  int<lower=0> N;
  array[N] int<lower=0, upper=1> y;
}
parameters {
  real<lower=0, upper=1> theta;
}
model {
  theta ~ beta(1, 1);
  y ~ bernoulli(theta);
}
generated quantities {
  array[N] real log_lik;
  for (n in 1:N) {
    log_lik[n] = bernoulli_lpmf(y[n] | theta);
  }
}
"

mod <- stan_model(code = bernoulli_model, model_name = "bernoulli")
data <- list(N = 10, y = c(1, 0, 1, 1, 0, 1, 0, 0, 1, 0))

fixed_param_model <- "
data {
  int<lower=0> N;
}
generated quantities {
  array[N] int<lower=0, upper=1> y_rep;
  for (n in 1:N) {
    y_rep[n] = bernoulli_rng(0.5);
  }
}
"

fixed_mod <- stan_model(code = fixed_param_model, model_name = "fixed_param")
fixed_data <- list(N = 5)
```

`stan_model()` returns an S3 object of class `"newstan_fit"` wrapping the
compiled model.

## Sampling

Use `sampling()` for MCMC. The result is a list with posterior draws, sampler
diagnostics, a Stan return code, and the normalized argument list.

```{r sampling}
fit <- sampling(
  mod,
  data,
  num_warmup = 20,
  num_samples = 20,
  num_chains = 2,
  seed = 123,
  verbose = FALSE
)

list(
  draws_class = class(fit$draws)[1],
  diagnostics_class = class(fit$diagnostics)[1],
  variables = posterior::variables(fit$draws),
  return_code = fit$return_code
)
```

The return value has:

- `draws`: a `posterior::draws_df` containing model parameter draws.
- `diagnostics`: a `posterior::draws_df` containing sampler diagnostics such as
  `accept_stat__`, `stepsize__`, `treedepth__`, `n_leapfrog__`,
  `divergent__`, and `energy__`.
- `return_code`: integer status code.
- `args`: named list of sampling arguments.

### Fixed Param

For models without parameters, or when you only want generated quantities, use
the fixed-parameter sampler.

```{r fixed-param}
fixed_fit <- sampling(
  fixed_mod,
  fixed_data,
  algorithm = "fixed_param",
  num_warmup = 0,
  num_samples = 10,
  seed = 123,
  verbose = FALSE
)

list(
  variables = posterior::variables(fixed_fit$draws),
  ndraws = posterior::ndraws(fixed_fit$draws),
  return_code = fixed_fit$return_code
)
```

### Supplying Inverse Metrics

`sampling()` accepts precomputed inverse metric values through `metric_file`.
This corresponds to Stan's `inv_metric` input.

For a diagonal metric, pass a numeric vector with one value per unconstrained
parameter. The Bernoulli model above has one unconstrained parameter.

```{r sampling-diag-metric}
diag_metric_fit <- sampling(
  mod,
  data,
  metric = "diag_e",
  metric_file = c(1.0),
  num_warmup = 20,
  num_samples = 20,
  seed = 123,
  verbose = FALSE
)

list(
  variables = posterior::variables(diag_metric_fit$draws),
  return_code = diag_metric_fit$return_code
)
```

For a dense metric, pass a square matrix. To provide one metric per chain, pass
a list whose length equals `num_chains`; a single vector or matrix is reused for
all chains.

```{r sampling-dense-metric}
dense_metric_fit <- sampling(
  mod,
  data,
  metric = "dense_e",
  metric_file = list(matrix(1.0, nrow = 1, ncol = 1)),
  num_chains = 1,
  num_warmup = 20,
  num_samples = 20,
  seed = 123,
  verbose = FALSE
)

list(
  variables = posterior::variables(dense_metric_fit$draws),
  return_code = dense_metric_fit$return_code
)
```

## Pathfinder

Use `pathfinder()` for an approximate posterior based on L-BFGS paths.

```{r pathfinder}
pf <- pathfinder(
  mod,
  data,
  max_lbfgs_iters = 100,
  num_paths = 1,
  num_draws = 50,
  seed = 123,
  verbose = FALSE
)

list(
  draws_class = class(pf$draws)[1],
  diagnostics_class = class(pf$diagnostics)[1],
  variables = posterior::variables(pf$draws),
  return_code = pf$return_code
)
```

The return value has `draws`, `diagnostics`, `return_code`, and `args`. If no
pathfinder diagnostic columns are returned, `diagnostics` is `NULL`.

## Variational Inference

Use `variational()` for ADVI with either a mean-field or full-rank Gaussian
family.

```{r variational}
vb <- variational(
  mod,
  data,
  algorithm = "meanfield",
  iter = 1000,
  output_samples = 50,
  seed = 123,
  verbose = FALSE
)

list(
  draws_class = class(vb$draws)[1],
  variables = posterior::variables(vb$draws),
  return_code = vb$return_code
)
```

The return value has `draws`, `return_code`, and `args`.

## Optimization

Use `optimizing()` to find a posterior mode or maximum likelihood estimate.

```{r optimizing}
opt <- optimizing(
  mod,
  data,
  algorithm = "lbfgs",
  seed = 123,
  verbose = FALSE
)

list(
  par = opt$par,
  value = opt$value,
  return_code = opt$return_code
)
```

The return value has `par`, `value`, `return_code`, and `args`.

## Laplace Approximation

Use `laplace()` to draw from a Laplace approximation around a mode. You can pass
the full `optimizing()` result or a named numeric vector of constrained
parameters.

```{r laplace}
lap <- laplace(
  mod,
  data,
  mode = opt,
  draws = 20,
  seed = 123,
  verbose = FALSE
)

list(
  draws_class = class(lap$draws)[1],
  variables = posterior::variables(lap$draws),
  return_code = lap$return_code
)
```

The return value has `draws`, `return_code`, and `args`.

## Generated Quantities

Use `generated_quantities()` with existing parameter draws and a model containing
a `generated quantities` block.

```{r generated-quantities}
gq <- generated_quantities(
  mod,
  data,
  fitted_params = fit$draws,
  seed = 123,
  verbose = FALSE
)

list(
  draws_class = class(gq$draws)[1],
  variables = posterior::variables(gq$draws),
  return_code = gq$return_code
)
```

The return value has `draws`, `return_code`, and `args`.

## Gradient Check

Use `gradient_check()` to compare autodiff gradients with finite-difference
approximations.

```{r gradient-check}
n_failed <- gradient_check(
  mod,
  data,
  seed = 123,
  verbose = FALSE
)

n_failed
```

`gradient_check()` returns an integer count of failed parameter gradient checks.

## Stan Compilation Helpers

`stanc()` compiles Stan code to C++ and returns the generated C++ source as a
single string.

```{r stanc}
cpp_code <- stanc(bernoulli_model)
substr(cpp_code, 1, 80)
```

`stan_model()` is usually the higher-level entry point: it calls `stanc()`,
compiles the generated C++, and returns the model object used by the services.
